---
title: "Untitled"
format: 
  html:
    toc: true # add table of contents
    toc-location: left # put toc on the left
    code-fold: true # all code is folded, can be clicked on to expand (only in html)
    theme: yeti # changes color theme I think
execute: 
  message: false
  warning: false
---

```{r libraries}

library(tidyverse)
library(here)
library(janitor)
library(ggeffects)
library(performance)
library(naniar)
library(flextable)
library(car)
library(broom)
library(corrplot)
library(AICcmodavg)
library(GGally)

```

Read in the data:

```{r reading-data}

plant <- read_csv(here("data", "knb-lter-hfr.109.18", "hf109-01-sarracenia.csv")) %>% 
  
  # make the column names cleaner
  clean_names() %>% 
  
  # selecting the columns of interest (only retain these columns)
  select(totmass, species, feedlevel, sla, chlorophyll, amass, num_lvs, num_phylls)

```

Visualize the missing data:

```{r missing-data-visualization}

gg_miss_var(plant)

```

Subsetting the data by dropping NAs:

```{r subset-drop-NA}

plant_subset <- plant %>% 
  
  # drops rows with NA values in the columns you specify
  # it's ok to remove NA's here because we don't know where NA's come from
  drop_na(sla, chlorophyll, amass, num_lvs, num_phylls) 

```

Create a correlation plot:

Example writing: To determine the relationships between numerical variables in our dataset, we calculated Pearsons r and visually represented correlation using a correlation plot.

```{r correlation-plot}

# calculate Pearson's R for numerical values only
plant_cor <- plant_subset %>% 
  
  # not including totmass because it's the y variable
  select(feedlevel:num_phylls) %>% 
  
  # make correlation matrix (plots variables against each other)
  # perfect correlation = 1
  cor(method = "pearson")

# creating a correlation plot (visualizes the correlation matrix we made with cor() function)
corrplot(plant_cor,
         # change shape of what's in the cells (ellipse pointing to right = positive relationship)
         method = "ellipse",
         addCoef.col = "black")

```

Create a plot of each variable compared against the others

```{r pairs-plot}

plant_subset %>% 
  
  # can include the species column now because pairs plots can include categorical variables too
  select(species:num_phylls) %>% 
  ggpairs()
```

Starting regression here:

(example) To determine how species and physiological characteristics predict biomass, we fit multiple linear models

```{r null-and-full-models}

# create a null model
# totmass ~ 1 means there is no predictor
null <- lm(totmass ~ 1, data = plant_subset)

# full model has all variables in it
full <- lm(totmass ~ species + feedlevel + sla + chlorophyll + amass + num_lvs + num_phylls, 
           data = plant_subset)

```

We visually assessed normality and homoscedasticity of residuals using diagnostic plots for the full model:

```{r full-diagnostics}

# set up a 2x2 grid to display diagnostic plots
par(mfrow = c(2,2))
plot(full)

# residuals look pretty normal because they are mostly along the QQ plot line
# looks heteroskedastic because even though line is pretty flat, residuals have a cone shape

```

We also tested for normality using the Shapiro-Wilk test (null hypothesis: variable of interest (i.e. the residuals) are normally distributed).

We tested for heteroskedasticity using the Breusch-Pagan test (null hypothesis: variable of interest has constant variance).

```{r}

check_normality(full)
# results: Non-normality of residuals detected (p < .001)
# this is very common for large datasets

check_heteroscedasticity(full)
# results: Heteroscedasticity (non-constant error variance) detected (p < .001).

# results: assumptions for multiple linear regression are not met

```

```{r}

# transform variables
# log function in R is natural log ln
full_log <- lm(log(totmass) ~ species + feedlevel + sla + chlorophyll + amass + num_lvs + num_phylls, data = plant_subset)

plot(full_log)
check_normality(full_log)
check_heteroscedasticity(full_log)

# results: residuals looks normal and homoscedastic now!
```







